\chapter{PURPOSE}
The underlying purpose of this policy is: 
	\begin{enumerate}
		\item To ensure that purchases and services contracted for the city are in the best interest of the public and to assure fair and equitable treatment of all persons who wish to or do conduct business with the city.
		\item To provide for the greatest possible economy in procurement activities for the city. 
		\item To foster effective broad-based competition within the free enterprise system to ensure that the city will receive the best possible service or product at the lowest possible price.
		\item To provide a systematic and uniform method of purchasing goods and services for the city. 
	\end{enumerate}

\chapter{DEFINITIONS}
\noindent
"Business" means any corporation, partnership, individual, sole proprietorship, joint stock company, joint venture, or any other private legal entity. 

\noindent
"Change order" means a written order signed by the city manager or department head, directing the contractor to suspend work or make changes, which the appropriate clauses of the contract authorize the city manager to order without the consent of the contractor or any written alteration in specifications, delivery point, rate of delivery, period of performance, price, quantity, or other provisions of any contract accomplished by mutual action of the parties to the contract. 

\noindent
"City" means Santa Clara City and all other reporting entities controlled by or financially dependent upon the city council. 

\noindent
 "City Manager" means the city manager of Santa Clara City. 

\noindent 
"Contract" means a written agreement for the delivery or disposal of supplies, services, or construction. 

\noindent 
"Invitation for bids" means all documents, whether attached or incorporated by reference, used for soliciting bids. 

\noindent
"Person" means any business, individual, union, committee, club, other organization, or group of individuals. 

\noindent
"Procurement" means buying, purchasing, renting, leasing, leasing with an option to purchase, or otherwise acquiring any supplies, services, or construction. 

\noindent
"Professional services contract" means a contract for services performed by an independent contractor in a professional capacity who produces a service predominantly of an intangible nature. These include, but are not limited to, the services of an attorney, physician, engineer, accountant, architectural consultant, dentist, artist, appraiser or photographer. 

\noindent
"Public utilities equipment and supplies" means equipment and materials purchased by the public works department or parks department that are used in the regular course of supplying public utilities or park and trail amenities, and includes but is not limited to such items as piping, irrigation equipment, meters, conduit, terminations, transformers, cable, wire, hoses, pad mounted equipment, and general construction materials. "Public utilities equipment and supplies" does not include any purchase that includes, in whole or in part, labor (not including freight or delivery), services, vehicles or machinery. 

\noindent
"Public works project" means the construction of a street, road, curb and gutter, sidewalk, power line or electrical generation or transmission facility; a public park, trail, or other recreational facility; a pipeline, culvert, dam, canal, or other system or facility for water, sewage, storm water, or flood control; or any other public infrastructure, system, or improvement constructed, owned, operated, or maintained by the City. 

\noindent
"Purchase" means the acquisition of goods (supplies, equipment, etc.) in a single transaction such that payment is made prior to receiving or upon receipt of the goods. 

\noindent
"Purchase description" means the words used in a solicitation to describe the supplies or services to be acquired, and includes specifications attached to or made a part of the solicitation. 

\noindent
"Request for proposals" means all documents, whether attached or incorporated by reference, used for soliciting proposals. 

\noindent
"Sealed bids" means written proposals from persons or entities offering to contract with or to sell to the city which are received in sealed envelopes to be opened after the deadline for receipt of bids as defined in an invitation to bid. 

\noindent
"Sole source" means goods or service for which there is only one source for the procurement item. Circumstances under which there is only one source for a procurement item may include: (1) where the most important consideration in obtaining a procurement item is the compatibility of equipment, technology, software, accessories, replacement parts, or service; (2) where a procurement item is needed for trial use or testing; (3) where transitional costs are unreasonable or cost prohibitive; or (4) where reasonably equivalent goods or services are not available from any other source. 

\chapter{POLICY}

\section{GENERAL POLICY}
\begin{enumerate}
	\item All city purchases and contracts for goods and services shall be subject to this chapter unless specifically exempted. 
	\item No contract or purchase shall be so arranged, fragmented or divided with the purpose or intent to circumvent this chapter. 
	\item No purchase shall be contracted for, or made, unless sufficient funds have been budgeted in the years in which the funds are to be expended. 
	\item Reasonable attempts shall be made to publicize anticipated purchases or contracts to known vendors, contractors and suppliers. 
	\item When it is advantageous to the city, annual or other recurring contracts for services and supplies regularly purchased should be approved. 
	\item All purchases and contracts must be approved by the city manager unless otherwise specified in this chapter. 
	\item All contracts for services shall be approved as to form by the city attorney. 
\end{enumerate}

\section{AUTHORITY OF CITY MANAGER}
The city manager or a person designated by the city manager shall be responsible for the following: 

\begin{enumerate}
	\item Ensure that all purchases for services comply with this policy. 
	\item Review and approve all purchases of the city. 
	\item Establish necessary additional procedures for the efficient and economical management of the contracting and purchasing functions authorized by this chapter. Such procedures shall be in writing and on file in the city recorder’s office as a public record. 
	\item Maintain accurate and sufficient records concerning all city purchases and contracts for services. 
	\item Maintain a list of contractors for public improvements and personal services who have made themselves known to the city and are interested in soliciting city business. 
	\item Make recommendations to the city council concerning amendments to this policy. 
	\item In the absence of the city manager, the mayor shall be designated and authorized to fulfill the duties of the city manager with respect to any contract, purchase or payment that requires immediate action or execution. The mayor shall promptly notify the city manager of any actions taken by the mayor. 
\end{enumerate}

\section{APPROVAL OF PURCHASES}
The following approvals are required for all purchases unless otherwise exempted in this policy:

\begin{enumerate}
	\item Any contract, purchase, purchase order, check request, or service request for nonprofessional services and supplies or building improvements in the total amount of \$1,000 or less may be made or approved by the city manager or department head, or an employee or supervisor authorized by the city manager or department head to make such purchase, purchase order, check request, or service request. 
	\item Any contract, purchase, purchase order, check request, or service request for nonprofessional services and supplies or building improvements in the total amount of more than \$1,000 but less than \$5,000 must be made or approved by the city manager or department head. 
	\item Any contract, purchase, purchase order, check request, or service request for nonprofessional services and supplies or building improvements in the total amount of more than \$5,000 but less than \$40,000 must be made or approved by the city manager. 
	\item Any contract, purchase, purchase order, check request, or service request for public utilities equipment and supplies in the total amount of more than \$5,000 but less than \$125,000 must be made or approved by both the public works director or parks director and the city manager. 
	\item The city manager may establish additional policies and controls which are consistent with the approval mechanism set forth in this chapter. 
	\item The following contracts and purchases must be approved by the city council: 
		\begin{enumerate}
			\item Any contract and purchase in the amount of \$40,000 or more, or any contract or purchase for public utilities equipment and supplies in the amount of \$125,000 or more. 
			\item Any contract awarded through the formal bidding process. 
			\item Any contract, purchase or payment over \$10,000 which is not anticipated in the current budget and is not funded by development fees (including connection fees or impact fees) or paid for by third parties. 
			\item Accumulated “change orders” which would increase a previously approved contract by \$5,000 or more, taking into consideration all previously approved change orders to the contract under consideration. 
		\end{enumerate}
	\item Credit cards issued in the city’s name may be made available for use by specific city employees.  Credit cards may be used for purchasing where there is some benefit to the city such as convenience, the ability to make recurring payments where required, or cash back to the city through a loyalty program.  Purchases using credit cards shall be subject to all the required limitations, requirements, and approvals of this chapter.  Each credit card statement shall be reviewed monthly by at least two persons, which shall include at least the city manager or city finance officer, and a department head, to protect against misuse or fraud.
\end{enumerate}

\pagebreak

\section{EXEMPTIONS}
The following transactions are exempt from the competitive bidding requirements of this chapter. The city manager shall determine whether or not a particular contract or purchase is exempt under the following criteria:

\begin{enumerate}
	\item When the provisions of this chapter would prevent the city from complying with the terms and conditions of any grant, gift, or bequest that is otherwise consistent with law.
	\item When any purchase or encumbrance is made with state or federal funds and the applicable state or federal laws or regulations are in conflict with this chapter to the extent that following the provisions of this chapter would jeopardize the use of those state or federal funds, such conflicting provisions of this chapter shall not apply and the city officials making such purchases shall follow the procedure required by the state or federal laws or regulations.
	\item Emergency contracts which require prompt execution of the contract because of an imminent threat to the safety or welfare of the public, or public or private property; circumstances which place the city or its officers and agents in a position of serious legal liability; or circumstances which are likely to cause the city to suffer financial harm or loss, the gravity of which clearly outweighs the benefits of competitive bidding in the usual manner. The city manager or mayor shall notify the city council of any emergency contract which would have normally required their approval at the next regularly scheduled city council meeting.
	\item Projects which are acquired, expanded, or improved under the Local Building Authority Act.
	\item Purchases from vendors or suppliers who are a sole source of the goods or services under procurement. If any sole source procurement exceeds \$40,000, notice of the sole source procurement shall be published in a newspaper of general circulation in the state or a newspaper of local circulation in the area at least 10 days before the public meeting to approve the purchase or contract by the city council.
	\item Purchases from state or local surplus property agencies or entities or purchases from entities under state-awarded equipment contracts. 
	\item Professional Service Contracts. The selection of personal service contracts shall be based on an evaluation of the services needed, the abilities of the contractors, the uniqueness of the service and the general performance of the contractor. The lowest quote need not necessarily be the successful contractor. Usually, emphasis will be placed on quality, with cost being the deciding factor when everything else is equal.
\end{enumerate}

\pagebreak

\section{PURCHASES NOT REQUIRING SEALED BIDS}
\begin{enumerate}
	\item Purchases and expenditures of less than \$1,000 shall not require bids or quotes of any type, but are otherwise subject to the general policies of this chapter. These purchases and expenditures are intended to include materials, supplies and services regularly purchased and consumed by the city such as office supplies, janitorial supplies, postage, food and beverages. Purchases shall not be artificially divided so as to constitute a purchase or expenditure under this section. 
	\item All purchases and expenditures of more than \$1,000 but less than \$5,000 shall require two documented price quotes unless the purchase or expenditure is exempt from the competitive bidding requirement of this chapter or requires a sealed bid. 
	\item All purchases and expenditures of more than \$5,000 shall require a minimum of three documented price quotes unless the purchase or expenditure is exempt from the competitive bidding requirement of this chapter or the purchase expenditure requires a sealed bid. The city manager may approve a purchase or expenditure without the required price quotes if the city manager determines that there are not two additional vendors or suppliers of such items. 
	\item Purchases made through the cooperative purchasing contracts administered by the State Division of Purchasing shall not require bids or quotes of any type. 
\end{enumerate}

\section{PURCHASES REQUIRING SEALED BIDS}
\begin{enumerate}
	\item Unless specifically exempt from the competitive bidding requirements of this chapter, whenever the total price of a contract for services, supplies or building improvements is estimated to be \$40,000 or more, or the total price of a contract for a public works project or public utilities equipment and supplies is estimated to be \$125,000 or more, an invitation for bids or a request for proposals shall be issued prior to acquisition. 
	\item Bid Specifications. Specifications for public contracts shall not expressly or implicitly require any product by any brand name or make, nor the product of any particular manufacturer or seller, unless the product is exempt by these regulations or by the directive of city council. 
	\item Advertising Requirements. An advertisement for bids is to be published at least twice in a newspaper of general circulation in the city, and in as many additional issues and publications as the city manager may determine, at least five days prior to the opening of bids. 
	\item All advertisements for bids shall state:
		\begin{enumerate}
			\item The date and time after which bids will not be accepted; 
			\item The date that prequalification applications must be filed, and the class or classes of work for which bidders must be prequalified if prequalification is a requirement; 
			\item The character of the work to be done or the materials or things to be purchased; 
			\item The office where the specifications for the work, material or things may be seen; 
			\item The name and title of the person designated for receipt of bids; 
			\item The type and amount of bid security if required; 
			\item The date, time and place that the bids will be publicly opened. 
		\end{enumerate}
\end{enumerate}

\pagebreak

\section{REQUIREMENTS FOR BIDS}
All bids made to the city shall be:

\begin{enumerate}
	\item In writing; 
	\item Filed with the city recorder, unless specifically stated in the bid requests; 
	\item Opened publicly by the city manager, department head or other designated person at the time designated in the advertisement and filed for public inspection; 
	\item Have the appropriate bid security attached, if required by the specifications for the project. 
\end{enumerate}

\section{AWARD OF CONTRACT}
After bids are opened, and a determination made that a contract be awarded, the award shall be made to the lowest responsible bidder. “Lowest responsible bidder” shall mean the lowest bidder who has substantially complied with all prescribed requirements, has submitted the bid in good faith, has a history of fully performing work at the bid price, and who has not been disqualified as set forth herein. 

The successful bidder shall promptly execute a formal contract and, if required, deliver a performance and payment bond to the city in a sum equal to the contract price, together with proof of appropriate insurance. Upon execution of the contract, bond, and insurance, bid security shall be returned. Failure to execute the contract, bond, or insurance shall be cause to forfeit the bid security. 

\section{REJECTION OF BIDS}
The city manager or the city council may reject any bid not in compliance with all prescribed requirements, and reject all bids if rejection of all bids is determined to be in the best interest of the city. 

\section{DISQUALIFICATION OF BIDDERS}
The city manager, upon investigation, may disqualify a bidder if: 

\begin{enumerate}
	\item The bidder does not have sufficient financial ability to perform the contract; 
	\item The bidder does not have equipment available to perform the contract; 
	\item The bidder does not have key personnel available, of sufficient experience, to perform the contract; 
	\item The person has previously breached contractual obligations with public or private agencies; or 
	\item The bidder fails to comply with the requests of an investigation by the city manager. 
\end{enumerate}

\pagebreak

\section{PREQUALIFICATION OF BIDDERS}
The city may require prequalification of bidders. 

\begin{enumerate}
	\item Upon establishment of the applicant’s qualifications, the city manager or department head shall issue a qualification statement. The statement shall inform the applicant of the project for which the qualification is valid, as well as any other conditions which may be imposed on the qualification. It shall advise the applicant to notify the city manager or department head promptly if there has been any substantial change of conditions or circumstances which would make any statement contained in the prequalification application no longer applicable or untrue. 
	\item If the city manager or department head does not qualify an applicant, written notice to the applicant is required stating the reasons the prequalification was denied and informing the applicant of the right to appeal the decision within five business days after receipt of the notice. Appeals shall be made to the city council. The city manager or department head may, upon discovering that a prequalified party is no longer qualified, revoke prequalification by sending notification of the revocation to the party subject to the revocation. The notice shall state the reason(s) for revocation, and that the revocation will be effective immediately. 
\end{enumerate}

\section{APPEAL PROCEDURES}
Any supplier, vendor, or contractor subject to an adverse procurement decision by the city, in violation of this chapter, may appeal that decision to the city council. 

\begin{enumerate}
	\item The complainant shall promptly file a written appeal letter, with the city recorder, within five working days from the time the alleged incident occurred. The letter of appeal shall state all relevant facts of the matter and the remedy sought. 
	\item Upon receipt of the notice of appeal, the city recorder shall forward to the city council the appeal notice, investigation of the matter and any other relevant information. 
	\item The city council shall conduct a hearing on the matter and provide the complainant an opportunity to be heard. A written decision shall be sent to the complainant. [Ord. O-17-2012]. 
\end{enumerate}

\pagebreak

\section{ACCOUNTS PAYABLE}
The city’s finance director will be responsible for establishing a system and process for the accurate and timely processing of all disbursements of city funds with emphasis on taking advantage of discounts when available and avoidance of late charges. 

\section{CANCELLATION AND REJECTION OF BIDS}
\begin{enumerate}
	\item An invitation for bids, a request for proposals, or other solicitation may be canceled, or any or all bids or proposals may be rejected, in whole or in part, as may be specified in the solicitation, when it is in the best interest of the city; the reasons for cancellation or rejection shall be made part of the contract file. 
	\item Bids, proposals, and contracts which create a conflict of interest under the Municipal Officers’ and Employees’ Ethics Act shall be rejected unless a legally required conflict disclosure has been made by the official or employee subject to the conflict. 
\end{enumerate}

\section{COST-PLUS-A-PERCENTAGE-OF-COST PROFITS CONTRACTS}
Subject to the limitations of this section, any type of contract which will promote the best interests of the city may be used; provided, that the use of a cost-plus-a-percentage-of-cost profit contract is prohibited unless specifically approved by the city council under special circumstances which clearly justify the use of such contracts because the scope of work or the nature of the work is to be closely supervised by the city and the city council so finds in writing. A cost-reimbursement contract may be used only when a determination is made in writing that such contract is likely to be less costly to the city than any other type or that it is impracticable to obtain the supplies, services, or construction required except under such a contract. 

\section{REQUIRED CONTRACT CLAUSES}
All contracts for work entered into by the city shall, whenever practicable, contain clauses permitting or requiring the following: 

\begin{enumerate}
	\item The unilateral right of the city to order, in writing, changes in the work within the scope of the contract and changes in the time of performance of the contract that do not alter the scope of the contract work. 
	\item Variations occurring between estimated quantities of work in a contract and actual quantities. 
	\item Suspension of work ordered by the city. 
	\item Requirements for performance bonds and labor and material payment bonds as required by law for construction projects. 
	\item Other provisions required for compliance with federal or state law. 
\end{enumerate}

All contract specifications shall seek to promote overall economy and best use for the purposes intended and encourage competition in satisfying the needs of the city and shall not be unduly restrictive. Where practical and reasonable, and within the scope of this chapter, Utah products shall be given preference as required by Utah law. 

\section{FINAL ACTION ON APPEALS}
The decision of the city council with respect to any appeal shall be the final action of the city with respect to that appeal. 

\section{CONFLICTS PROHIBITED}
No person involved in making procurement decisions may have personal investments or derive any income from any business entity which may create a conflict between their private interests and their public duties.  In the event of a conflict, the conflicted person shall recuse themselves from the particular procurement decision or decisions creating the conflict by notifying the city manager (or, in the event the city manager is the conflicted party, by notifying the mayor) of the conflict, and shall not participate in that particular decision or decisions. 

\section{EFFECTIVE DATE}
The foregoing ordinance shall be effective immediately upon adoption and posting. 

\textbf{ADOPTED} and approved by a duly constituted quorum of the City Council of the City of Santa Clara, Utah, this 24th day of April, 2018. 